Low-Complexity Adaptive Sonar Imaging

### 2019-10-23 LCAforActiveSonarImaging/submission/final/Buskenes_Jo_Inge-Low_Complexity_Adaptive_Sonar_Imaging.tex
2019-10-22

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEcompsoctitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
\IEEEcompsoctitleabstractindextext{%
\begin{abstract}
% 
% \begin{itemize}
% \item Dynamic weights $\Rightarrow$ greater contrast/resolution potential
% \item But:
% \begin{itemize}
% \item Active system $\Rightarrow$ noise and signal correlated $\Rightarrow$ adaptive beamformers no longer robust.
% \item Subarray averaging decorrelate the noise, but in the process we sacrifice resolution.
% \item Computationally complex O($M^3$) due to inversion of a spatial covariance matrix. LCA is only O($MW$). Even if Capon subspace/beamspace versions exist, they remain more complex than the LCA, and is not as ideally suited for implementation of SIMD hardware (GPUs, DSPs, ...).
% \end{itemize}
% \item LCA is inherently robust, is ``parameter free'' once setup with a well designed window set, and can often be implemented with minor modifications to already existing DAS beamformers.
% 
% The angular resolution and contrast in active sonar images depend on the beamformer's ability to receive signals from directions of interest, while suppressing noise and interference emanating from other directions. For sonar arrays, this is achieved by applying weights to the array channels.
% 
% While classical beamformers use predefined windows, adaptive beamformers estimate the optimal window by analytical evaluation of the data. The minimum variance (MV) beamformer, for instance, calculates the set of weights that minimises the variance of the beamformer's output. 
% 
We have studied the low complexity adaptive (LCA) beamformer in active sonar imaging. LCA can be viewed as either a simplification of the minimum variance distortionless response (MVDR) beamformer, or as an adaptive extension to the delay and sum (DAS) beamformer. While both LCA and MVDR attempt to minimize the power of noise and interference in the image, MVDR achieves this by computing optimal array weights from the spatial statistics of the wavefield, while LCA selects the best performing weights out of a predefined set.

To build confidence in the LCA method we show that a robust MVDR implementation typically creates weights spanning between the rectangular and Hamming window function. We let LCA select from a set of Kaiser windows with responses in this span, and add some steered variations of each. We limit the steering to roughly half the \minus{}3\,dB width of the window's amplitude response. Using experimental data from the Kongsberg Maritime HISAS1030 sonar we find that LCA and MVDR produce nearly identical images of large scenes, both being superior to DAS. On point targets LCA is able to double the resolution compared to DAS, or provide half that of MVDR. This performance is achieved with a total of 6 windows; the rectangular window and the Kaiser window with $\beta=5$, in an unsteered version, and versions that are left and right steered to the steering limit. Slightly smoother images are produced if the window count is increased to 15, but past this we observe minimal difference. Finally we show that LCA works just as well if Kaiser windows are substituted with trigonometric ones.

@@

% An attactive trait of the LCA is its low computational complexity; while the MV beamformer is of O($M^3$), the LCA method is of O($MW$), with $M$ being the number of channels and $W$ the number of windows. We made the LCA perform like the MV method using a well designed set of 30 windows. Hence, unless the array is very small, the proposed method will perform like the MV beamformer or better, and at a fraction of the computational cost.

@@

% (Optional) Add some extra info on cover page of peer review papers:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi

@@

% - Nature and scope
% - Past work
% - Methods
% - Brief summary results
% - Brief conclusions

% The promise is already demonstrated in nature.
% 
% The sonar systems of the future 
% 
% Future sonar systems 
% 
% Great promise in adaptivity
% 
% The greatest untapped 
% 
% The sonars of the future might benefit greatly from adaptivity.
% 
% 
% Future sonars will likely be

% To make a sonar truly perform better, we need to make it smarter. 
% 
% \IEEEPARstart{F}{uture} sonar systems will merit a much more sophisticated level of adaptivity. 
% 
% \IEEEPARstart{T}{he} next generation sonar systems will be adaptive.
% 
% Making sonars adaptive is the way to take them into the future. 
% 
% The virtue of future sonar systems will be sophisticated adaptivity. 
% 
% Intelligent
% 
% Reactive
% 
% Observe, process, adapt, act. 
% 
% Adaptivity - sense, process, respond
% 
% Faster, smarter
% 
% The best sonars in nature are adaptive. 
% Nature allures us to be this bold because

\IEEEPARstart{T}{he} best sonars in existence is perhaps found in nature. Bats, for instance, use a high frequency sonar to detect, identify and track prey with amazing precision in caves amongst a myriad of other bats. No man made sonar can currenly match this feat, much due to the challenge of processing and adapting to such a complex environment fast enough. This is why human made sonars have little to no adaptivity on transmit. However, several methods have been investigated to adaptively post process the received data. Perhaps the best studied method for reconstructing a sonar image out of the data is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}. It adapts the sonar array's spatial response to minimize the influence of noise and interference in the final image.

In many cases MVDR can improve the contrast and resolution of active sonar images compared to conventional static methods~\cite{Blomberg2013,Blomberg2012a,Lo2004}. However, while the computational complexity of conventional beamformers are linear with the number of channels, O($M$), MVDR is at O($M^3$). This is because MVDR relies on estimating and inverting a covariance matrix. The estimation step dominates the computation in sonar systems with less than 32 channels, but here the computational complexity can be reduced and the implementation accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014}. For larger systems the inversion step dominates. This can be dealt with by approximating the full covariance matrix with a small one using space reduction techniques such as beamspace processing~\cite{Asen2013}, or the closely related principal component analysis (PCA) method~\cite{Kim2014}. Another alternative is to assume spatial stationarity to form more easily invertible Toepliz matrices~\cite{Asl2012}. However, the mentioned methods are still relatively slow compared to DAS.

A much faster and simpler alternative to MVDR is the Low Complexity Adaptive (LCA) beamformer. Based on an idea by Vignon~\cite{Vignon2008}, it was first introduced by Synnev\aa{}g \emph{et al}. in clinical medical ultrasound imaging~\cite{Synnevag2011}, who demonstrated its ability to obtain very similar image quality to MVDR in systems with focused transmit beams. LCA applies a set of predefined and static windows, or apodization weights, and selects the one offering the best noise suppression. It relies on the same optimization criterion as MVDR, thus may be considered a version of MVDR with a discrete and static window solution space. Alternatively, it may be viewed as a multi-apodization technique. One such method is described by Stankwitz \emph{et al}. in radar imaging~\cite{Stankwitz1995}, where the best out of 2 or 3 windows are selected. However, LCA is different in that it typically selects from a larger pool of windows and allows phase-steered variations of them. This gives LCA an adaptive freedom that lies somewhere in between that of traditional multi-apodization techniques and MVDR. How similarly LCA performs compared to MVDR depends on how well the predefined windows represent the constrained solution space of MVDR.

% While using weights that span a bigger space gives LCA greater freedom, not The bigger the space the weights span, the greater the adaptive freedom, but only the solutions that are valid in the constrained solution space can make a difference.

% Instead of adaptively computing the weights, it adaptively selects them.

Synnev\aa{}g \emph{et al}. suggested using a window set comprised of rectangular, Kaiser (or Kaiser-Harris) and inverted Kaiser functions. In total he made LCA choose from 12 unique windows, 6 of which were phase-shifted versions of a fairly wide Kaiser window, and one were rectangular. Phase-shifting the windows has the effect of slighly adjusting the angle the array is steered towards. Synnev\aa{}g's composition of windows was motivated by the desire to use wide windows with maximum sidelobe suppression near the receiver, and narrow windows at greater depths for maximum sensitivity and depth penetration. While achieving good results, it was not made clear how many windows LCA needs, or how they ideally should be steered. Also, it is not apparent how LCA performs in sonar systems where the transmit beam is unfocused.

In this study we expand our previous work of applying LCA to active sonar imaging~\cite{Buskenes2011}. We use variations of the parametric Kaiser window, but also compare it to the trigonometric window function. We extend Synnev\aa{}g's work by steering all our windows and not just some, and demonstrate how LCA's performance is affected by the type and number of windows included. Our results are obtained using experimental data from the 32 element Kongsberg Maritime HISAS1030 sonar, which operates at 100\,kHz with an \minus{}3\,dB element opening angle of 23$^\circ$. Although this system was designed as synthetic aperture sonar (SAS) where an image is synthesized from several pings of data, we will only create images out of single pings (sectorscan imaging) in this study.

Our results show that when MVDR is modified to work in active sonar imaging, it computes weights with spatial responses that tend to be symmetric and steered within a fraction of the window's \minus{}3\,dB resolution. When letting LCA choose from Kaiser windows with similar responses, we obtain images with improved noise suppression and resolution. The resolution gain is made predictable by limiting the maximum steering to a fraction of the window's \minus{}3\,dB width. Overall, LCA produces images that are similar to MVDR, but with a resolution in between that of DAS and MVDR. LCA is also inherently robust, easy to implement, and fairly easy to understand.

This article is outlined as follows: In Section \ref{sec:beamforming} we offer a gentle introduction to beamforming. Then we move on to adaptive beamforming and describe the MVDR and LCA methods in \ref{sec:mvdr} and \ref{sec:lca}, respectively. We use LCA with the Kaiser window which we describe in \ref{sec:lca_kaiser_windows}, explain how we steer it in \ref{sec:lca_steering}, and by how much in \ref{sec:lca_steering_bounds}. Finally we add some remarks on using the trigonometric window instead of Kaiser in \ref{sec:lca_trigonometric}, and of of sampling considerations in \ref{sec:lca_oversampling}. Results and discussion are provided in Section \ref{sec:results_discussion}, where we study MVDR's windows to find suitable LCA ones in \ref{sec:results_lca_window_function}, assess the range of window types and steering in \ref{sec:results_window_parameters}, determine the number of windows needed in \ref{sec:results_database_size}, and add a note on computational complexity in \ref{sec:results_complexity}. Section \ref{sec:conclusion} provides a conclusion.

% This article is outlined as follows: In Section \ref{methods}, we introduce the concept of adaptive beamforming and provide details on the MVDR and LCA method. Then, in \ref{maptogpu}, we investigate the complexity issue, discuss means for reducing arithmetic complexity, and detail an implementation that makes efficient use of the GPU's parallel resources. The final design is assessed in Sections \ref{images_and_benchmarks} and \ref{discussion}, where we provide benchmarks, comparisons with similar CPU implementations, and measures of how efficiently our implementation makes use of the GPU's resources.

% - How much oversampling
% - How many parameters, how great a span.
% 
% - Experiementl data, Holmengraa/cross
% 
% - Note on efficiency.
% 
% 
% From our work with MVDR it 
% 
% 
% Traditionally beamformers were static, but as processors get more powerful so with the emergence of more processing power

% 
% \section{Methods}
% 
% At heart of the image formation process is a technique called beamforming, which creates a focus point at the pixel of interest. It does this by applying a suitable set of delays and weights to the array elements, such that signals emanating from the pixel location are emphasized constructively, while noise and interference from other directions sum destructively.
% 
% In active sonar array imaging, an acoustic wave is transmitted and the received echoes are usually recorded using an array of sensors. Each of these may be independently delayed and weighted such that signals emanating from directions of interest are added constructively, while noise and interference from other directions add destructively. Additionally, a window may be applied to the sensor channels to further adjust the array's spatial response. This process is known as beamforming.% The process of combining the signals from each sensor is known as beamforming.
% 
% 
% 
% 
% % Traditional beamformers such as Delay-and-Sum (DAS) apply predefined windows to all incoming data. However, due to the non-stationary nature of sonar data, the optimal window for any given time instant will generally differ from the next. This is where adaptive beamformers thrive, because they compute the optimal window coefficients for the data at each time instant. The choice of optimisation criteria is what mainly differentiates the various adaptive beamformers. The Minimum Variance (MV) beamformer, for instance, selects the set of weights that minimizes the beamformer output power for any given time instant~\cite{cap69}.
% 
% 
% \section{Background}
% 
% The MV beamformer computes the window coefficients by estimating and inverting a spatial covariance matrix. For improved estimation accuracy the data is averaged in space or time, or both. In addition, a certain amount of diagonal loading is often added to the covariance matrix before inversion. These actions, while leading to statistical robustness, tend to deteriorate the beamformer's performance. Also, the inversion of the covariance matrix has a computational complexity of O($M^3$), where $M$ is the number of elements. The computational burden makes the MV beamformer for large arrays less attractive, and sometimes even infeasible.
% 
% Based on the proposed method by Synnevåg~\cite{syn11}, we have implemented the LCA beamformer that keeps the minimum variance optimisation criterion but reduces the solution space to a discrete set of predefined windows. This reduces the computational complexity to O($MW$), where $W$ is the number of windows in the set. We use the Kaiser window function because it allows us to design a wide range of windows with different mainlobe widths and sidelobe suppression by adjusting the tradeoff parameter $\beta$~\cite{kai66}. In addition we apply steering to each of the windows and constrain the window design to ensure unit gain in the look direction.
% 
% The following will show that the proposed method performs similarly to or better than the MV method.

@@

% \begin{figure*}[t]%\centering%
% \includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds_lca_windows_beta.pdf}%
% \caption{$\beta$-values of the windows LCA selects for each image pixel. The underlying image is the one shown in \Fig{oversampling_mosaic_bounds}, with the same parameters. The location of the anchor and cut line is marked in red. Note how LCA prefers a rectangular window ($\beta=0$) on the anchor.}\label{oversampling_mosaic_bounds_beta}
% \end{figure*} 
% 
% \begin{figure*}[t]%\centering%
% \includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds_lca_windows_phi.pdf}%
% \caption{$\beta$-values of the windows LCA selects for each image pixel. The underlying image is the one shown in \Fig{oversampling_mosaic_bounds}, with the same parameters. The location of the anchor and cut line is marked in red. Note how LCA prefers a rectangular window ($\beta=0$) on the anchor.}\label{oversampling_mosaic_bounds_beta}
% \end{figure*}

% % \setcounter{dbltopnumber}{1

% \begin{figure*}[t]%\centering%
% \includegraphics[width=\textwidth]{gfx/beamformer_comparison_cross.pdf}%
% \caption{}\label{beamformer_comparison_cross}
% \end{figure*}
% 
% \begin{figure*}[t]%\centering%
% \includegraphics[width=\textwidth]{gfx/beamformer_comparison_holmengraa.pdf}%
% \caption{}\label{beamformer_comparison_holmengraa}
% \end{figure*}

@@

If the pixels are correlated laterally, we can include a weighted combination of these as well to improve the variance estimation:
%
\begin{align}
s_z^2 = \frac{1}{N_x N_k} \sumb{\nmat{\scriptstyle x'=\\[-2mm]\scriptstyle x-X}}{x+X} \sumb{\nmat{\scriptstyle n'=\\[-2mm]\scriptstyle n-K}}{n+K} \omega[x',n']\big| z[x',n'] \big|^2
\end{align}
%
where $N_x = 2X+1$ is the number of azimuth lines to average over, and $\omega[x',n']$ is a normalized 2 dimensional weight function. Since we oversample slightly laterally when we delay each pixel (see Section \ref{sec:lca_oversampling}), we thought the lateral correlation to be sufficient to benefit from this. However, we observed no visual improvement from applying this technique, and decided not to use it.
% 
% \begin{itemize}
% \item \emph{Window function}
% \item \emph{Parameter boundaries}
% \item \emph{Number of windows needed}
% \item \emph{Oversampling need}
% \item \emph{Image quality}
% \end{itemize}

@@

Since LCA operates with the same optimization criterion as MVDR, we will be using approximately a factor 8 lateral oversampling for both beamformers. The images appear more visually pleasing and detailed up to 8. We consider an absolute minimum to be 2, due to the non-linear nature of delaying pixels that isn't strictly in the far field, and of displaying the absolute value of the pixels on a decibel scale. In \mediaI{}\footnote{For media location refer to Appendix \ref{app:source_code}.} we visualize the effect of changing the lateral oversampling factor on a set of LCA images. Upon display the images are all bilinearly upinterpolated to the same size. 
% 
% \todo{Need $M$ in the image sector. Why???}

% Idea to steer within a fraction of the rectangular window's mainlobe.

% \begin{align}
% d = \frac{\pi\,\theta_\text{open}\,\lambda}{M\,d\,O_s}
% \end{align}

% See section \ref{kaiser_windows}.

% Narrow ones:
% - Best coherent gain, lowest equivalent noise bandwidth, and most sensitive to steering meaning not needing to steer as much for some effect.
% 
% 
% Beamforming:
% - Maximize resolution
% - Minimize noise/error, both coherent and incoherent
% - Minimize signal cancellation
% 
% Gauss: Miminized time-bandwidth product
%   - 
%   
% Rectangular
%   - 
% Dolph-Cheb: Minimize mainlobe width given sidelobe suppression
% Kaiser: For a restricted energy and time duration, maximize energy in the band of frequencies W.
% 
% Adaptive windows perform best in detection of closely separated angles of significantly amplitudes.

% Adaptive beamformers usually compute the window coefficients by estimating and inverting a spatial covariance matrix. There are two inherent problems to this procedure. First, for improved estimation accuracy the data is either averaged in space or time, and the significance of the covariance matrix can be reduced using a regularisation parameter~\cite{Carl Inge}. These are all attempts to constrain the beamformer to ensure the window coefficience are not over-adapted to the data. Second, inverting the covariance matrix has a computational complexity of O($M^3$)~\cite{Carl Inge}. For larger arrays the computational burden becomes significant, and the question arises whether this processing power can be better utilised. 
% 
%

% 
% This selection is based on observations of the solutions found by the MV beamformer. We obtained good results in our experiments by letting the LCA beamformer select among 5 Kaiser windows with uniformly distributed $\beta$'s in the range $[0.05, 0.5]$ and a rectangular window. Each window was steered in 5 different directions, uniformly distributed within 80\% of the mainlobe width of the rectangular window, which is the most narrow. This adds to a total of 30 windows.
% 
% To make the choice of window less susceptible to pixel value uncertainties, the LCA beamformer was set to apply the most frequently selected window in an 11 pixel range region to the center pixel in that region. In physical terms, this is a region of approximately 20\,cm. The MV beamformer estimated the covariance matrix as described in~\cite{syn07} by averaging over 16 subarrays and 11 range pixels, and applying 3\% diagonal loading.

@@

The results will be presented in the following order: In section \ref{sec:results_lca_window_function} we discuss typical window responses computed by MVDR, and hypothesize that these can be mimicked by Kaiser windows. In section \ref{sec:results_window_parameters} we determine sensible boundaries for the Kaiser parameter $\beta$ and window steering $\phi$. In section \ref{sec:results_database_size} we discuss how many window variations are needed for LCA to perform well.

@@

Assuming that a robust MVDR is the reference method we want LCA to perform similarly to, it seems sensible to create a window database for LCA with spatial responses similar to the ones that MVDR computes. We study this in \Fig{mvdr_selected_windows}, where we present the typical amplitude and phase responses that MVDR creates for shadow, speckle and highlight regions of the Holmengraa scene.

Observe from \Fig{mvdr_selected_windows} that MVDR seems to prefer symmetric window responses, even if it is free to choose non-symmetric ones. The symmetry is most predominant in shadow and speckle regions, while in highlight the windows are approximately symmetric within the illuminated sector of the seafloor. We have observed this symmetry to be a side-effect of the averaging steps needed to build the sample covariance matrix, which are required for MVDR to operate in near snapshot mode in an active system. The peak sidelobe levels of the MVDR windows are mostly between \minus{}13\,dB and \minus{}43\,dB. This corresponds to that of an unsteered rectangular and Hamming window, or an unsteered Kaiser window with $\beta=0$ and $\beta=2$, respectively.

From this we hypothesize that a good window database LCA can be made using a varied  set of Kaiser windows. If they span a suitable range of $\beta$-values, and some steering variations are applied to each, we should have responses that resemble those in \Fig{mvdr_selected_windows}. The Kaiser window is easy to compute, is fairly insensitive to coefficient inaccuracies, and can create windows with shapes spanning from rectangular to Gaussian. It is also optimal in the sense of having its peak power concentration near the steering angle.

Unlike the method described by Synnev\aa{}g~\cite{Synnevag2011} we do not let LCA choose from inverted Kaiser windows. These have a mainlobe width narrower than that of the rectangular window, but at the expense of a very poor white noise gain. We have found that these windows hardly ever get used on experimental data, in particular steered versions of them. We can infer the same from \Fig{mvdr_selected_windows} by noting that the maximum sidelobe level rarely exceeds the rectangular window level of \minus{}13\,dB.

\mediaII{}\footnote{For media location refer to Appendix \ref{app:source_code}.} animates how \Fig{mvdr_selected_windows} changes as a function of the MVDR subarray length $L$ and temporal averaging $K$. At $L=1$ the window responses are rectangular in all areas in the image. At $L=2$ we observe window responses with slight amplitude variations in the highlight region, but with the phase being 0$^\circ$ in the illuminated seafloor sector. Already at $L=2$ the MVDR is able to greatly suppress noise. This is a common observation; adding a little flexibility to adapt to the scene has a dramatic effect, but further allowing full flexibility is much less significant. The media file also shows that MVDR can be run with subarray sizes $L\in[M/2,5M/8]$, but only with temporal averaging $K=1$ or above.

@@

To determine a suitable range for the Kaiser parameter $\beta$ and steering $\phi$, we constructed large and equally sized window databases containing Kaiser windows with varied upper boundaries for $\beta$ and $\phi$. The lower boundaries were chosen as $\beta=0$ and $\phi=0^\circ$, which includes the rectangularly weighted DAS image. The resulting images are shown in \Fig{oversampling_mosaic_bounds}. In each image we computed the lateral distance $\Delta x$ between the \minus{}3\,dB points of the anchor, with the reference $\Delta x_\text{ref}$ being the rectangularly weighted DAS image. A fourth order polynomial fit was used in this computation. This measure is also commonly called the full width half maximum (FWHM). In the case of a imaging a point scatter, it is closely related to the resolution of of the system. For a rectangular window, for instance, the system resolution is approximately $\Delta \approx \text{FWHM}/0.89$~\cite{Harris1978}.

>>>>

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEcompsoctitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
\IEEEcompsoctitleabstractindextext{%
\begin{abstract}
% 
% 
We have studied the low complexity adaptive (LCA) beamformer in active sonar imaging. LCA can be viewed as either a simplification of the minimum variance distortionless response (MVDR) beamformer, or as an adaptive extension to the delay and sum (DAS) beamformer. While both LCA and MVDR attempt to minimize the power of noise and interference in the image, MVDR achieves this by computing optimal array weights from the spatial statistics of the wavefield, while LCA selects the best performing weights out of a predefined set.

To build confidence in the LCA method we show that a robust MVDR implementation typically creates weight sets with shapes spanning between a rectangular and Hamming window function. We let LCA select from a set of Kaiser windows with responses in this span, and add some steered variations of each. We limit the steering to roughly half the \minus{}3\,dB width of the window's amplitude response. Using experimental data from the Kongsberg Maritime HISAS1030 sonar we find that LCA and MVDR produce nearly identical images of large scenes, both being superior to DAS. On point targets LCA is able to double the resolution compared to DAS, or provide half that of MVDR. This performance is achieved with a total of 6 windows; the rectangular window and the Kaiser window with $\beta=5$, in an unsteered version, and versions that are left and right steered to the steering limit. Slightly smoother images are produced if the window count is increased to 15, but past this we observe minimal difference. Finally we show that LCA works just as well if Kaiser windows are substituted with trigonometric ones.

@@

@@

@@

\IEEEPARstart{T}{he} best sonars in existence is perhaps found in nature. Bats, for instance, use a high frequency sonar to detect, identify and track prey with amazing precision in caves amongst a myriad of other bats. No man made sonar can currenly match this feat, much due to the challenge of processing and adapting to such a complex environment fast enough. This is why human made sonars have little to no adaptivity on transmit. However, several methods have been investigated to adaptively process the received acustic data. Perhaps the best studied method for reconstructing a sonar image out of such data is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}. It adapts the sonar array's spatial response to minimize the influence of noise and interference in the final image.

In many cases MVDR can improve the contrast and resolution of active sonar images compared to conventional static methods~\cite{Blomberg2013,Blomberg2012a,Lo2004}. However, while the computational complexity of conventional beamformers are linear with the number of channels, O($M$), MVDR is at O($M^3$). This is because MVDR relies on estimating and inverting a covariance matrix. The estimation step dominates the computation in sonar systems with less than 32 channels, but here the computational complexity can be reduced and the implementation accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014}. For larger systems the inversion step dominates. This can be dealt with by approximating the full covariance matrix with a small one using space reduction techniques such as beamspace processing~\cite{Asen2013}, or the closely related principal component analysis (PCA)~\cite{Kim2014}. Another alternative is to assume spatial stationarity to form more easily invertible Toepliz matrices~\cite{Asl2012}. However, the mentioned methods are still relatively slow compared to DAS.

A much faster and simpler alternative to MVDR is the Low Complexity Adaptive (LCA) beamformer. Based on an idea by Vignon~\cite{Vignon2008}, it was first introduced by Synnev\aa{}g \emph{et al}. in clinical medical ultrasound imaging~\cite{Synnevag2011}, who demonstrated its ability to obtain very similar image quality to MVDR in systems with focused transmit beams. LCA applies a set of predefined and static windows, or apodization weights, and selects the one offering the best noise suppression. It relies on the same optimization criterion as MVDR, thus may be considered a version of MVDR with a discrete and static window solution space. Alternatively, it may be viewed as a multi-apodization technique. One such method is described by Stankwitz \emph{et al}. in radar imaging~\cite{Stankwitz1995}, where the best out of 2 or 3 windows is selected. However, LCA differs in that it typically selects from a larger pool of windows and allows phase-steered variations of them. Phase-shifting the windows has the effect of slighly adjusting the angle the array is steered towards. This gives LCA an adaptive freedom that lies somewhere in between that of traditional multi-apodization techniques and MVDR. How similarly LCA performs compared to MVDR depends on how well the predefined windows represent the constrained solution space of MVDR.

Synnev\aa{}g \emph{et al}. suggested using a window set comprised of rectangular, Kaiser (or Kaiser-Harris) and inverted Kaiser functions. In total he let LCA choose from 12 unique windows, 6 of which were phase-shifted versions of a fairly wide Kaiser window, and one were rectangular. Synnev\aa{}g's composition of windows was motivated by the desire to use wide windows with maximum sidelobe suppression near the receiver, and narrow windows at greater depths for maximum sensitivity and depth penetration. While he achieved good results, it was not made clear how many windows LCA needs, or how they ideally should be steered. Also, it is not apparent how LCA performs in sonar systems where the transmit beam is unfocused.

In this study we expand our previous work of applying LCA to active sonar imaging~\cite{Buskenes2011}. We use variations of the parametric Kaiser window, but also compare it to the trigonometric window function. We extend Synnev\aa{}g's work by steering all our windows and not just some, and demonstrate how LCA's performance is affected by the type and number of windows included. Our results are obtained using experimental data from the 32 element Kongsberg Maritime HISAS1030 sonar, which operates at 100\,kHz with an \minus{}3\,dB element opening angle of 23$^\circ$. Although this system was designed as a synthetic aperture sonar (SAS) where an image is synthesized from several pings of data, we will only create images out of single pings (sectorscan imaging) in this study.

Our results show that when MVDR is robustified to work in active sonar imaging, it computes weights with spatial amplitude responses that tend to be symmetric and steered within a fraction of their \minus{}3\,dB widths. When letting LCA choose from Kaiser windows with similar responses, we obtain images with improved noise suppression and resolution. The resolution gain is made predictable by limiting the maximum steering to a fraction of the \minus{}3\,dB width of each window's spatial amplitude response. Overall, LCA produces images that are similar to MVDR, but with a resolution in between that of DAS and MVDR. LCA is also inherently robust, easy to implement, and fairly easy to understand.

This article is outlined as follows: In Section \ref{sec:beamforming} we offer a gentle introduction to beamforming. Then we move on to adaptive beamforming and describe the MVDR and LCA methods in \ref{sec:mvdr} and \ref{sec:lca}, respectively. We use LCA with the Kaiser window which we describe in \ref{sec:lca_kaiser_windows}, explain how we steer it in \ref{sec:lca_steering}, and by how much in \ref{sec:lca_steering_bounds}. Finally we add some remarks on using the trigonometric window instead of Kaiser in \ref{sec:lca_trigonometric}, and of of sampling considerations in \ref{sec:lca_oversampling}. Results and discussion are provided in Section \ref{sec:results_discussion}, where we study MVDR's windows to find suitable LCA ones in \ref{sec:results_lca_window_function}, assess the range of window types and steering in \ref{sec:results_window_parameters}, determine the number of windows needed in \ref{sec:results_database_size}, and add a note on computational complexity in \ref{sec:results_complexity}. Finally, Section \ref{sec:conclusion} provides a conclusion.

@@

@@

If the pixels are correlated laterally, we can include a weighted combination of these to improve the variance estimation:
%
\begin{align}
s_z^2 = \frac{1}{N_x N_k} \sumb{\nmat{\scriptstyle x'=\\[-2mm]\scriptstyle x-X}}{x+X} \sumb{\nmat{\scriptstyle n'=\\[-2mm]\scriptstyle n-K}}{n+K} \omega[x',n']\big| z[x',n'] \big|^2
\end{align}
%
where $N_x = 2X+1$ is the number of azimuth lines to average over, and $\omega[x',n']$ is a normalized 2 dimensional weight function. Since we oversample slightly laterally when we delay each pixel (see Section \ref{sec:lca_oversampling}), we thought the lateral correlation to be sufficient to benefit from this. However, we observed no visual improvement from applying this technique, and decided not to use it.
%

@@

Since LCA operates with the same optimization criterion as MVDR, we will be using approximately a factor 8 lateral oversampling for both beamformers. The images appear more visually pleasing and detailed up to 8. We consider an absolute minimum to be 2, due to the non-linear nature of delaying pixels that isn't strictly in the far field, and of displaying the absolute value of the pixels on a decibel scale. In \mediaI{}\footnote{For media location refer to Appendix \ref{app:source_code}.} we visualize the effect of changing the lateral oversampling factor on a set of LCA images. Upon display the images were all bilinearly upinterpolated to the same size.

@@

The results will be presented in the following order: In Section \ref{sec:results_lca_window_function} we discuss typical window responses computed by MVDR, and hypothesize that these can be mimicked by Kaiser windows. In Section \ref{sec:results_window_parameters} we determine sensible boundaries for the Kaiser parameter $\beta$ and window steering $\phi$. In Section \ref{sec:results_database_size} we discuss how many window variations are needed for LCA to perform well.

@@

Assuming that a robust MVDR is the reference method we want LCA to perform similarly to, it seems sensible to create a window database for LCA with spatial responses similar to the ones that MVDR computes. We study this in \Fig{mvdr_selected_windows}, where we present the typical amplitude and phase responses for the windows that MVDR creates in shadow, speckle and highlight regions of the Holmengraa scene.

Observe from \Fig{mvdr_selected_windows} that MVDR seems to prefer symmetric window responses, even if it is free to choose non-symmetric ones. The symmetry is most predominant in shadow and speckle regions, while in highlight the windows are approximately symmetric within the illuminated sector of the seafloor. We have observed this symmetry to be a side-effect of the averaging steps needed to build the sample covariance matrix, which are required for MVDR to operate with very little to no temporal sample support in an active system. The peak sidelobe levels of the MVDR windows are mostly between \minus{}13\,dB and \minus{}43\,dB. This corresponds to that of an unsteered rectangular and Hamming window, or an unsteered Kaiser window with $\beta=0$ and $\beta=2$, respectively.

From this we hypothesize that a good window database LCA can be made using a varied  set of Kaiser windows. If they span a suitable range of $\beta$-values, and some steering variations are applied to each, we should have responses that resemble those in \Fig{mvdr_selected_windows}. The Kaiser window is easy to compute, is fairly insensitive to coefficient inaccuracies, and span shapes from rectangular to Gaussian. It is also optimal in the sense of having its peak power concentration near the steering angle.

Unlike the method described by Synnev\aa{}g~\cite{Synnevag2011} we do not let LCA choose from inverted Kaiser windows. These have a mainlobe width narrower than that of the rectangular window, but at the expense of very poor white noise gain. We have found that these windows hardly ever get used on experimental data, in particular steered versions of them. We can infer the same from \Fig{mvdr_selected_windows} by noting that the maximum sidelobe level rarely exceeds the rectangular window level of \minus{}13\,dB.

\mediaII{}\footnote{For media location refer to Appendix \ref{app:source_code}.} animates how \Fig{mvdr_selected_windows} changes as a function of the MVDR subarray length $L$ and temporal averaging $K$. At $L=1$ the window responses are rectangular in all areas in the image. At $L=2$ we observe window responses with slight amplitude variations in the highlight region, but with the phase being 0$^\circ$ in the illuminated seafloor sector. Already at $L=2$ the MVDR is able to greatly suppress noise. This is a common observation; adding a little flexibility to adapt to the scene has a dramatic effect, but allowing full flexibility is much less significant. The media file also shows that MVDR can be run with subarray sizes $L\in[M/2,5M/8]$, but only with temporal averaging $K=1$ or above.

@@

To determine a suitable range for the Kaiser parameter $\beta$ and phase steering $\phi$, we constructed large and equally sized window databases containing Kaiser windows with varied upper boundaries for $\beta$ and $\phi$. The lower boundaries were chosen as $\beta=0$ and $\phi=0^\circ$, which includes the rectangularly weighted DAS image. The resulting images are shown in \Fig{oversampling_mosaic_bounds}. In each image we computed the lateral distance $\Delta x$ between the \minus{}3\,dB points of the anchor, with the reference $\Delta x_\text{ref}$ being the rectangularly weighted DAS image. A fourth order polynomial fit was used in this computation. The measure $\Delta x$ is also commonly called the full width half maximum (FWHM). In the case of a imaging a point scatter, it is closely related to the resolution of the system. For a rectangular window, for instance, the system resolution is approximately $\delta x \approx \Delta x/0.89$~\cite{Harris1978}.

The LCA images in \Fig{oversampling_mosaic_bounds} demonstrate the effect of adjusting the upper bound of $\beta$ and $\phi$. As suggested by \Fig{mvdr_selected_windows} we first attempted to use $\beta\in[0,2]$. While this significantly improved sidelobe suppression, a slight further improvement was observed up to $\beta\in[0,5]$. This can likely be explained by the need to compensate for the increased sidelobe levels caused by steering. No noticeable difference were observed for higher values than $\beta=5$. The images also demonstrate that increasing the upper bound for steering $\phi$ improves the resolution of the strong scatterers. For this sonar the Rayleigh resolution is $\delta x\approx0.0125$ radians (0.72$^\circ$), so a point scatterer imaged at 19\,m range with a rectangular DAS would have a lateral $\Delta x\approx0.0125\cdot19\,\text{m}\cdot0.89=21\,\text{cm}$. This is close to the FWHM of the anchor in the DAS image in \Fig{oversampling_mosaic_bounds}, which means that the acoustic fingerprint of the anchor is similar to a point source. This further implies that the anchor reflections are specular and that we view its rounded side. As we increase steering to either 50\% or 100\% of the \minus{}3\,dB width, the FWHM drops to 73\% or 53\% to that of the rectangular window, respectively.
